Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-07-21 threshold/draft.Rnw
a few typos
Notation fixes

Let $A=U\Sigma V^T$ be the singular value decomposition of $A$, where
$U\in\mathcal{R}^{m\times k}$, $U^TU = I$, $V\in\mathcal{R}^{n\times k}$,
$V^TV=I$, and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$. 
For later convenience, we write $s=\mathrm{diag}(\Sigma)$ to be the
diagonal vector of the matrix $\Sigma$.
Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
$v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
V\Sigma^2 V^T$ is a symmetric eigenvalue decomposition of the correlation
matrix with eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$. The span
of the columns of $V$ form an orthonormal basis of the range of the correlation
matrix $A^TA$.

@@

Including more terms from Equation \ref{vdist} increases our ability to prune
pairs of vectors outside the correlation threshold. For example, including
only one more term finds that only 116 out of 418 pairs of adjacent vectors
in the example shown in Figure \ref{fig1} satisfy
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 \le 2(1-t),
\]
where exponentiation is applied element-wise to the vectors.
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including five terms reduces the total to only 11 pairs of vectors.
In general, using $p\le k$ terms boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}) \le 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values,
and $(D^1PV_{1:p})^2$ denotes element-wise exponentiation.

@@

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$, correlation threshold $t$, maximum
subspace dimension $p<<n$.
\begin{enumerate}
\item Compute the truncated SVD $AV_{1:p} = U_{1:p}s_{1:p}$, $V_{1:p} = [v_1, v_2, \ldots, v_p]$.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p} \le 2(1-t)
\right\},
\]
where $e_{\tilde{j}}\in\mathcal{R}^n$ consists of $1$ in position $\tilde{j}$ and
zeros otherwise.
\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and go to
step 1. Continue to increase $p$ in this way until the number of candidate
pairs stops decreasing.
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.

>>>>

Let $A=U\Sigma V^T$ be the singular value decomposition of $A$, where
$U\in\mathcal{R}^{m\times k}$, $U^TU = I$, $V\in\mathcal{R}^{n\times k}$,
$V^TV=I$, and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  For
later convenience, we write $s=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$ to be
the vector of the first $p\le k$ singular values along the diagonal of
$\Sigma$.  Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
$v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
V\Sigma^2 V^T$ is a symmetric eigenvalue decomposition of the correlation
matrix with eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$. The span
of the columns of $V$ form an orthonormal basis of the range of the correlation
matrix $A^TA$.

@@

Including more terms from Equation \ref{vdist} increases our ability to prune
pairs of vectors outside the correlation threshold. For example, including
only one more term finds that only 116 out of 418 pairs of adjacent vectors
in the example shown in Figure \ref{fig1} satisfy
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 \le 2(1-t),
\]
where exponentiation is applied element-wise to the vectors.
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including five terms reduces the total to only 11 pairs of vectors.
In general, using $p\le k$ terms boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values,
and exponentiation is applied element-wise.

@@

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$, correlation threshold $t$, maximum
subspace dimension $p<<n$.
\begin{enumerate}
\item Compute the truncated SVD $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where $V_{1:p} = [v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\},
\]
where $e_{\tilde{j}}\in\mathcal{R}^n$ consists of $1$ in position $\tilde{j}$ and
zeros otherwise.
\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and go to
step 1. Continue to increase $p$ in this way until the number of candidate
pairs stops decreasing.
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.
